---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Day 1: Report Repair

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)

library(tidyverse)
library(scales)
```

```{r}
input <- read_file("../input/01 - Report Repair - Input.txt") %>%
  str_split("\\n") %>%
  unlist() %>%
  as.numeric()
```

### Puzzle 1

Find the two entries that sum to 2020 and then multiply those two numbers together.

```{r}
get_answer <- . %>%
  pivot_longer(contains("input"), names_to = "input_num", values_to = "input") %>%
  select(input, sum, product) %>%
  distinct(input, sum, product)

solution1 <- crossing(input1 = input, input2 = input) %>%
  mutate(sum = input1 + input2, 
         product = input1 * input2) %>%
  filter(sum == 2020) %>%
  get_answer()
```

The two numbers that add to 2020 are **`r paste(solution1 %>% pull(input), collapse = ", ")`**. The solution is therefore **`r formattable::comma(prod(solution1 %>% pull(input)), digits = 0)`**.

### Puzzle 2

What is the product of the three entries that sum to 2020?

```{r}
solution2 <- crossing(input1 = input, input2 = input, input3 = input) %>%
  mutate(sum = input1 + input2 + input3, 
         product = input1 * input2 * input3) %>%
  filter(sum == 2020) %>%
  get_answer()
```

The two numbers that add to 2020 are **`r paste(solution2 %>% pull(input), collapse = ", ")`**. The solution is therefore **`r formattable::comma(prod(solution2 %>% pull(input)), digits = 0)`**