---
output: html_document
editor_options: 
  chunk_output_type: console
---

## Day 7: Handy Haversacks

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)

library(tidyverse)
library(scales)
```

```{r}
input <- read_file("input/07 - Handy Haversacks.txt") %>%
  str_split("\\n") %>%
  unlist() %>%
  tibble(input = .)

cleaned_input <- input %>%
  extract(input, c("outer", "inner"), regex = "([\\w\\s]+)bags contain (.*)?\\.", remove = FALSE) %>%
  mutate(outer = str_trim(outer),
         inner = map(inner, ~ str_extract_all(., pattern = "[\\d+|no][\\w\\s]+bags?")[[1]])) %>%
  unnest(inner) %>%
  mutate(inner = str_trim(str_replace(inner, "bags?", ""))) %>%
  separate(inner, c("num", "inner"), sep = " ", remove = FALSE, extra = "merge") %>%
  mutate(num = as.integer(ifelse(num == "no", 0, num)), 
         inner = na_if(inner, "other")) %>%
  select(outer_bag = outer, inner_bag = inner, can_contain = num)
```

### Puzzle 1

Using the example, what I hope to accomplish is to turn this into a graph problem. With a directed graph, I can find the paths from each vertex to "shiny gold". Once I have all the paths, the bags that can fit the shiny gold bag will be all the unique vertices that were traversed through.

```{r}
library(ggraph)
library(igraph)
library(widyr)
```

```{r, include=FALSE}

example_text1 <- paste(
"light red bags contain 1 bright white bag, 2 muted yellow bags.",
"dark orange bags contain 3 bright white bags, 4 muted yellow bags.",
"bright white bags contain 1 shiny gold bag.",
"muted yellow bags contain 2 shiny gold bags, 9 faded blue bags.",
"shiny gold bags contain 1 dark olive bag, 2 vibrant plum bags.",
"dark olive bags contain 3 faded blue bags, 4 dotted black bags.",
"vibrant plum bags contain 5 faded blue bags, 6 dotted black bags.",
"faded blue bags contain no other bags.",
"dotted black bags contain no other bags.",
sep = "\n")

dex1 <- example_text1 %>%
  str_split("\\n") %>%
  unlist() %>%
  tibble(input = .) %>%
  extract(input, c("outer", "inner"), regex = "([\\w\\s]+)bags contain (.*)?\\.", 
          remove = FALSE) %>%
  mutate(outer = str_trim(outer),
         inner = map(inner, ~ str_extract_all(., pattern = "[\\d+|no][\\w\\s]+bags?")[[1]])) %>%
  unnest(inner) %>%
  mutate(inner = str_trim(str_replace(inner, "bags?", ""))) %>%
  separate(inner, c("num", "inner"), sep = " ", remove = FALSE, extra = "merge") %>%
  mutate(num = as.integer(ifelse(num == "no", 0, num)), 
         inner = na_if(inner, "other")) %>%
  select(outer_bag = outer, inner_bag = inner, can_contain = num)

gex1 <- dex1 %>%
  filter(!is.na(inner_bag)) %>%
  graph_from_data_frame()

ggraph(gex1) + 
  geom_edge_link(aes(
      start_cap = label_rect(node1.name), 
      end_cap = label_rect(node2.name)
    ), arrow = arrow(type = "closed", length = unit(3, 'mm'))
  ) + 
  geom_node_text(aes(label = name)) +
  theme_graph() + 
  labs(title = "Example Problem 1")
```


Drawing the full problem would be very messy, so I use tidy verbs to obtain the paths.

```{r}
clean_path <- function(p) {
  tibble(node = p) %>%
    mutate(path = row_number(), 
           node = map(node, ~ names(.))) %>%
    unnest(node) %>%
    group_by(path) %>%
    mutate(step = row_number()) %>%
    ungroup()
}

my_bag <- "shiny gold"

g <- cleaned_input %>%
  select(outer_bag, inner_bag) %>%
  filter(!is.na(inner_bag)) %>%
  as.matrix() %>%
  graph_from_edgelist(directed = TRUE)

solution1 <- tibble(start = names(V(g))) %>%
  mutate(path = map(start, ~ all_simple_paths(g, ., to = my_bag, mode = "out")),
         path = map(path, clean_path)) %>%
  unnest(path)
  
n_bags <- solution1 %>% count(start) %>% nrow()
```

The number of bag colors that can eventually contain at least one shiny gold bag is **`r n_bags`**.

### Puzzle 2

```{r}
example_text2 <- paste(
  "shiny gold bags contain 2 dark red bags",
  "dark red bags contain 2 dark orange bags, 1 dark green bag",
  "dark orange bags contain 2 dark yellow bags.",
  "dark yellow bags contain 2 dark green bags.",
  "dark green bags contain 2 dark blue bags.",
  "dark blue bags contain 2 dark violet bags.",
  "dark violet bags contain no other bags.",
sep = "\n")

dex2 <- example_text2 %>%
  str_split("\\n") %>%
  unlist() %>%
  tibble(input = .) %>%
  extract(input, c("outer", "inner"), regex = "([\\w\\s]+)bags contain (.*)", 
          remove = FALSE) %>%
  mutate(outer = str_trim(outer),
         inner = map(inner, ~ str_extract_all(., pattern = "[\\d+|no][\\w\\s]+bags?")[[1]])) %>%
  unnest(inner) %>%
  mutate(inner = str_trim(str_replace(inner, "bags?", ""))) %>%
  separate(inner, c("num", "inner"), sep = " ", remove = FALSE, extra = "merge") %>%
  mutate(num = as.integer(ifelse(num == "no", 0, num)), 
         inner = na_if(inner, "other")) %>%
  select(outer_bag = outer, inner_bag = inner, can_contain = num)

gex2 <- dex2 %>%
  filter(!is.na(inner_bag)) %>%
  graph_from_data_frame()

ggraph(gex2) + 
  geom_edge_link(aes(
      start_cap = label_rect(node1.name), 
      end_cap = label_rect(node2.name)
    ), arrow = arrow(type = "closed", length = unit(3, 'mm'))
  ) + 
  geom_node_text(aes(label = name)) +
  theme_graph() + 
  labs(title = "Example Problem 2")
```

```{r}
end_nodes <- dex2 %>% filter(can_contain == 0) %>% pull(outer_bag)

tibble(end = end_nodes) %>%
  mutate(path = map(end, ~ all_simple_paths(gex2, my_bag, to = ., mode = "out")),
         path = map(path, clean_path)) %>%
  unnest(path) %>%
  mutate(outer_bag = node, inner_bag = lead(outer_bag)) %>%
  left_join(dex2, by = c("outer_bag", "inner_bag")) %>%
  group_by(end, path) %>%
  filter(step != max(step)) %>%
  mutate(num_bags = cumprod(can_contain)) %>% 
  ungroup() %>%
  count(end, step, outer_bag, inner_bag, num_bags)
  summarise(tot_bags = sum(num_bags), .groups = "drop")

n_bags2 <- solution2 %>% pull(tot_bags) %>% sum()
n_bags2
```

